Joint estimation of ancestral states and phylogenetic model parameters is inconsistent

### 2017-12-02 joint_inf.tex
errors in estimation

An alternative approach \cite{Sagulenko2017-jo} does away with the marginalization and directly estimates the maximum likelihood parameters of the fully-observed likelihood in \eqref{eq:full_likelihood}.
This is known in statistics as a profile likelihood \cite{Murphy2000-ry}, which exists here because $\alphabet$ is a finite set:
\begin{equation}
\label{eq:profile_likelihood}
L_\nCols'(\tau, t; \fullAlignment) = \max_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates).
\end{equation}
We will use $\hat{\fullAncestralStates}$ to denote an estimate for $\fullAncestralStates$ obtained by maximization of \eqref{eq:profile_likelihood}.
One can then estimate a topology and branch fidelities using this profile likelihood:
\begin{equation}
\label{eq:profile_likelihood_topology_bl}
(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \ L_\nCols'(\tau, t; \fullAlignment).
\end{equation}
In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
For this reason, in practice, the joint ML strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \cite{Sagulenko2017-jo}.

@@

Marginal inference performs as expected, where $\hat{w}$ is equal to $y^*$ in the limit data regardless of the value of $x^*$.
For joint inference, the estimates for $\hat{w}$ when $x^*$ and $y^*$ are both large look unbiased, with $\hat{w}$ increasing as $y^*$ increases, though the region where $\hat{w}$ is estimated to be one---indicating no change along the interior branch---is similar to that in the restricted case.
Comparing Figs.~\ref{fig:bl-inconsistency} and~\ref{fig:bl-general-inconsistency} shows that the region of inconsistency in the general case is similar to that of the restricted case.
This region encompasses the majority of the parameter space; even given the correct topology and performing our best possible optimization, we have many situations where we will estimate the interior branch parameter to be one.
As a positive control, Fig.~\ref{fig:bl-general-marginal} shows the result of optimizing \eqref{eq:marginal_likelihood} using the same procedure.
Here we obtain the correct $\hat{w}\rightarrow y^*$.

>>>>

An alternative approach \cite{Sagulenko2017-jo} does away with the marginalization and directly estimates the maximum likelihood parameters of the fully-observed likelihood in \eqref{eq:full_likelihood}.
This is known in statistics as a profile likelihood \cite{Murphy2000-ry}, which exists here because $\alphabet$ is a finite set:
\begin{equation}
\label{eq:profile_likelihood}
L_\nCols'(\tau, t; \fullAlignment) = \max_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\tau, t; \fullAlignment, \fullAncestralStates).
\end{equation}
We will use $\hat{\fullAncestralStates}$ to denote an estimate for $\fullAncestralStates$ obtained by maximizing \eqref{eq:profile_likelihood}.
One can then estimate a topology and branch fidelities using this profile likelihood:
\begin{equation}
\label{eq:profile_likelihood_topology_bl}
(\hat{\tau}, \hat{t}) = \argmax_{\tau, t} \ L_\nCols'(\tau, t; \fullAlignment).
\end{equation}
In general, the functional form of \eqref{eq:profile_likelihood} is determined by inequalities that depend on the unknown $(\tau,t)$.
For this reason, in practice, the joint ML strategy estimates $\hat{\fullAncestralStates}$ for a fixed $(\tau,t)$, then $(\hat{\tau},\hat{t})$ given $\hat{\fullAncestralStates}$, maximizing each of these conditional objectives until convergence \cite{Sagulenko2017-jo}.

@@

Marginal inference performs as expected, where $\hat{w}$ is equal to $y^*$ in the limit data regardless of the value of $x^*$.
For joint inference, the estimates for $\hat{w}$ when $x^*$ and $y^*$ are both large look unbiased, with $\hat{w}$ increasing as $y^*$ increases, though the region where $\hat{w}$ is estimated to be one---indicating no change along the interior branch---is similar to that in the restricted case.
Comparing Figs.~\ref{fig:bl-inconsistency} and~\ref{fig:bl-general-inconsistency} shows that the region of inconsistency in the general case is similar to that of the restricted case.
This region encompasses the majority of the parameter space; even given the correct topology and performing our best possible optimization, we have many situations where we will estimate the interior branch parameter to be one.
As a positive control, Fig.~\ref{fig:bl-general-marginal} shows the result of optimizing \eqref{eq:marginal_likelihood} using the same procedure.
Here we obtain the correct $\hat{w}\rightarrow y^*$.
To understand the quality of each fit, we report the range of $\hat{w}-y^*$ where $\hat{w}\neq 1$.
For joint inference, the errors range from $[-7\times 10^{-3}, 8\times 10^{-2}]$ and for marginal inference, $[-8\times 10^{-8}, 5\times 10^{-7}]$ showing that, even in cases where joint inference does not estimate $\hat{w}$ to be exactly one, it still fails to achieve a low error from truth.
